# (PART) R functionality {-} 

# R functions
In previous sections we've seen R **functions** that are used on objects to perform some activity. Functions seen so far include:

- `class()` and `is.*()` functions for checking datatypes
- `as.*` for converting to datatypes
- `length()` and `names()` for metadata
- `head()` and `tail()` for getting a small amount of elements from an object
- `ncol()`, `nrow()`, `colnames()`, and `rownames()` for getting data.frame metadata
- `Sys.Date()` and `Sys.time()` for getting current date-time values

There are a huge range of functions out there, whether available in R straight away, or from adding extra functionality.

Understanding how functions work and being able to use them correctly will help you learn, and use R effectively.

## Using a function
A function does some computation on an object. The use of a function consists of:

1. A function's name
2. Parentheses
3. 0 or more inputs

Each input is provided to an **argument** or parameter within a function.

These arguments have names, although you don't often need to provide the names.

You can find out what arguments a function takes by using the code completion and it's help snippet, or by searching for the function in the Rstudio Help tab.

When you're inside the brackets of a function you can get the list of available arguments and auto-complete them.

```{block 04_functions-1, type="readerTODO"}
Try getting the code completion to work with the function `tail()`.
```

## Examining functions
One of the niftiest things about R is being able to see the code for a function. You can examine how many functions work by just typing their name without any parentheses.

```{r 04_functions-2}
Sys.Date
```

The first line(s) show how the arguments are specified. Subsequent lines show the code and the final lines starting with `<` can be mostly ignored.


## Function input patterns
Functions tend to conform to certain patterns of inputs. 

### No inputs
Some functions don't require the user to provide info and so they don't have any arguments. `Sys.Date()` and similar functions do not need user input because the functions provide information about the system.

```{r 04_functions-3}
Sys.Date
```

Looking at the function definition, we can see that there are no arguments specified in the first line.

### Single inputs
Other functions only have a single allowed input. `length()` returns the length of an object so it only allows you to provide it with an object.

```{r 04_functions-4}
length
```

We can see in this definition^[the code looks a bit odd - this is because it's specified a bit differently to most functions, but fear not! In R, there are some different ways of writing code. The normal way is called **S3** but functions that are designed to work with properties of objects like length use a different system called **S4**. For the most part you'll rarely need to dig into S4 code as most R functionality is built in S3 and will allow you to check out it's code easily.] that the function takes the argument `x`.

### Many inputs
Some functions have multiple inputs, although not all of them are necessarily **mandatory**. `head()` and `tail()` have been used so far with only a single input but they take an optional argument as to how many elements should be returned.

```{r 04_functions-5}
head(letters)
head(letters, 2)
```

The `rnorm()` function allows us to generate a vector of values from a normal distribution. We can tell it how many values we need (`n`), and we can optionally provide the mean (`mean`) and standard deviation (`sd`) to describe the Normal curve that values should be selected from.

```{r 04_functions-6}
rnorm
```

Looking at how `rnorm` is specified we can see that we're expected to provide `n`, but `mean` and `sd` are given values of 0 and 1 respectively by default. 

```{r 04_functions-7}
rnorm(n=5)
rnorm(n = 5, mean = 10, sd = 2)
```

### Unlimited inputs
Other functions can take an unlimited amount of input values. Functions like `sum()` will sum the values from a number of objects.

```{r 04_functions-8}
sum(1:3, 1:9, pi)
```

The ellipsis (`...`) is used to denote when the user can provide any number of values.

```{r 04_functions-9}
sum
```

## Naming arguments
Every input provided to a function is associated with an argument. 

Each argument must have a name. Even functions that allow unlimited inputs assign these inputs to a name. Behind the scenes, they get put into a list object and the list gets called `...` (or ellipsis).

There are some typical names for arguments that take your data object. These include:

- `x`
- `data`
- `.data`
- `df`

You don't usually have to provide the argument names, just put things in the relevant places in the function. Sometimes though, you *will* need to use argument names. 

Here are my rules of thumb for knowing when you need to name names:

1. You're using the arguments in an order that is different from the function author's intended order (you might be skipping some arguments as the default values are fine or you might just prefer a different order)
2. The arguments you want to specify show up after the `...` in a function's argument list
3. You want to give a specific name to a value in a `...` argument


We can provide names^[Note that you don't have to type the full name as R will attempt to match up values, but doing that can and does get a lot of R users into trouble so I don't recommend it.] for clarity or so we can use arguments out of order if we prefer to.

```{r 04_functions-10}
rnorm(n = 5, mean = 10, sd = 2)
rnorm(mean = 10, sd = 2, n = 5)
```

A common behaviour change that you'll need to work with is how missing (`NA`) values get handled. Functions that allow you change this behaviour, usually have an argument called things like `na.rm`, `na.omit`, and `na.action`.

```{r 04_functions-11}
sum(1:5, NA)
sum(1:5, NA, na.rm = TRUE)
```

In the `sum()` example, I used the `na.rm` argument's name. This is because otherwise the `TRUE` would be considered part of the values being passed for summing. Without the name, the value gets considered as part of the `...`.

```{r 04_functions-12}
sum(1:5, NA, TRUE)
```


A function will sometimes have `...` at the end of it's list of arguments when it utilises other functions and those have optional / default values.

For instance the `predict()` function allows us to take a model we've built and apply it to some new data. 

It works for many different types of model and these different models expect different types of inputs. Some models expect data.frames, others expect time series data, etc. 

There's lots of potential variations, the only thing that is mandatory is the model object. 

```{r 04_functions-13}
predict
```

The `predict()` function then determines what type of model object you've provided it and passes the model, and any other values you provided, to the relevant function, returning back the results.

```{r 04_functions-14, warning=FALSE}
linearMod<-lm(Sepal.Length~., data=iris)
logisticMod<-glm(Species~., data=iris, family=binomial)

predict(linearMod, iris[1,])
predict(logisticMod, iris[1,])
```

## Summary
R uses functions as the means of performing operations^[Indeed, even operators like `+` are actually functions behind the scenes]. 

Functions can take 0 or more arguments. All arguments may be mandatory, but some can be optional or even undefined.

You can use argument names to provide arguments in different orders to that defined by the function author or to provide them in the case where an ellipsis (`...`) is used in a function.

## R functions Exercises
Using what you've learned about investigating the components of functions...

1. Use `pmin()` to find the smallest values element-wise of the three vector `1:51`, `25:75`, `30:-20`
2. Use `paste()` to combine the upper case letters into a single string with `", "` between each letter
3. Use `list.files()` to see what files are in your current directory. Return the fully qualified names not just the file names
4. View the code for `ncol()` and work out how the number of columns is being determined


