# (PART) Portable code {-}
# Making functions

```{r Basic_R_Functions-1, include=FALSE}
library(tidyverse)
```


## Anatomy of a function
### Functions
Functions are bundled code that has some or all of the optional components:

- Inputs
- Outputs
- Side effects

### Inputs
Inputs are arguments that you pass into a function.

```{r Basic_R_Functions-2}
inputsEx<-function(x,y){}
inputsEx(1,2)
```

### Outputs
The function's output is the object the function returns. It's usually a transformation of the input(s) in some way.

```{r Basic_R_Functions-3}
outputsEx<-function(x){
  result<-x*2
  return(result)
}
outputsEx(3)
```

### Side effects
These are things that happen but are not part of the output. Generally you should avoid doing this so that everything that happens in the fufnction, stays in the function.

```{r Basic_R_Functions-4}
sideffectsEx<-function(x){
  result<-x*2
  plot(x,result)
  return(result)
}
sideffectsEx(1:10)
```

### tidyverse
We can also write functions in the tidyverse paradigm for compact code.
```{r Basic_R_Functions-5}
outputsEx<-rlang::as_function(~ .*2)
outputsEx(3)
```

### Exercises
1. Write a function that calculates the area of circles given the radius as the input argument

## Arguments
### Mandatory arguments
Arguments that don't get provided with a default value in the definition are mandatory.

```{r Basic_R_Functions-6}
argsEx<-function(x, y=2){
  return(x*y)
}
argsEx(2)
argsEx(2,3)
```

### Arguments with a defined set of values
```{r Basic_R_Functions-7, error=TRUE}
argsSubsetEx<-function(x){
  x<-match.arg(x,c("red","green","blue"))
  return(x)
}
argsSubsetEx("red")
argsSubsetEx("Red")
```

### Arguments to provide a function
```{r Basic_R_Functions-8}
argsFunctionEx<-function(func,x){
  func<-match.fun(func)
  return(func(x))
}
argsFunctionEx("mean", 1:10)
```

### Arbitrary arguments
The ellipsis (`...`) is used to allow the function user to put in any number of arguments - these might be used by the core code, or they could be passed into functions called within the function.

```{r Basic_R_Functions-9}
ellipsisEx1<-function(...){
  args<-list(...)
  return(args)
}
ellipsisEx1(LETTERS,1:5)
```

### Arbitrary arguments
```{r Basic_R_Functions-10}
ellipsisEx2<-function(x, ...){
  smallest<-pmin(...)
  return(x*smallest)
}
ellipsisEx2(5,1,2,3)
ellipsisEx2(8,3:10,10:3)
```

### Exercises
1. Write a function that takes some strings and will upper or lower case them. Only allow upcase or lower casing.
2. Write a function that returns the number of arbitrary arguments passed into itself

## Documentation
### Roxygen
R has a documentation system for functions and packages. When we write a function, especially a complex one, we should add some documentation to make it easier to understand.

Get a roxygen skeleton by being in the function definition and pressing ctrl+shift+alt+r

### Basic documentation
```{r Basic_R_Functions-11}
#' Title
#'
#' @param func 
#' @param x 
#'
#' @return
#' @export
#'
#' @examples
argsFunctionEx<-function(func,x){
  func<-match.fun(func)
  return(func(x))
}
```

### Basic documentation
```{r Basic_R_Functions-12}
#' Perform any function on any single input
#'
#' @param func The name of a function as a string
#' @param x The input to be provided in the first position to the function
#'
#' @return The evaluated results of func(x)
#' @export
#'
#' @examples
#' argsFunctionEx(tolower, LETTERS)
argsFunctionEx<-function(func,x){
  func<-match.fun(func)
  return(func(x))
}
```

### Exercises
1. Document the functions you've written so far

## Error handling

### Providing information to the user
There are different "severities" of text you can provide back to the user of a function:

- message
- warning
- error

### Messages
Designed to provide information that might be helpful but doesn't say there's a problem.

```{r Basic_R_Functions-13}
messageEx<-function(x){
  message("Life is good!")
  x
}
messageEx(5)
```

### Warnings
Designed to let the user know that the function output might not be what they expect.

```{r Basic_R_Functions-14}
warningsEx<-function(x){
  if(sum(is.na(x))) warning("Some NAs in the data - results might not match expectations")
  mean<- sum(x, na.rm=TRUE) / length(x)
  return(mean)
}
warningsEx(c(1:10, NA))
```

### Errors
When something means the function cannot return a result, you should make the function return an informative error.

```{r Basic_R_Functions-15, error=TRUE}
stopEx<-function(x, y){
  if(all.equal(y,0)) stop("y is 0")
  return(x/y)
}
stopEx(1,2)
stopEx(1,0)
```

### Handling unexpected inputs
We can check the types or ranges of input values by applying assertions that have to be passed in order for the function to execute.

```{r Basic_R_Functions-16, error=TRUE}
stopifnotEx<-function(x){
  stopifnot(is.numeric(x), x>10, x<20)
  return(x)
}
stopifnotEx("A")
stopifnotEx(10)
```

### Code that can fail
Some code could error for one reason or another. You might want the function to return a success any way or do something different in the event of an error.

```{r Basic_R_Functions-17}
tryEx<-function(x,y){
  tryCatch({x/y},
           error=function(e){
             message("Check your y")
             })
}
tryEx(1,"A")
tryEx(1,2)
```

### Exercises
1. Write a function that adds together the square root of two arguments. 
2. Add a message that return the result rounded to 2 decimal places
2. Add a warning if either argumenet is NA
3. Add a check step that will throw an error if either argument is not numeric
