```{r Basic_R_IO-1, include=FALSE, message=FALSE}
library(dplyr)
library(DBI)
library(odbc)
library(readr)
library(readxl)
library(openxlsx)
library(httr)
```


# CSVs
## Reading CSVs
```{r Basic_R_IO-2}
survey<-readr::read_csv("../data/sample.csv")
```

```{r Basic_R_IO-3}
survey<-readr::read_csv("../data/sample.csv",
                        col_names = c("entry","name","icecream","pet", 
                                      "weekday","floors","yearborn"),
                        skip=1)
```

## Writing CSVs
```{r Basic_R_IO-4}
readr::write_csv(head(survey),"sample2.csv")
```

```{r Basic_R_IO-5, echo=FALSE, results='hide'}
file.remove("sample2.csv")
```

## Exercises
1. Load sample.csv with sensible names and use `col_*` functions to convert the date and the weekday columns to sensible data types as you load them.
2. Write a random sample of the loaded data to a new file

# Spreadsheets

## Reading spreadsheets
```{r Basic_R_IO-6}
survey<-readxl::read_excel("../data/sample.xlsx")
```

```{r Basic_R_IO-7}
readxl::excel_sheets("../data/sample.xlsx")
```


```{r Basic_R_IO-8}
survey<-readxl::read_excel("../data/sample.xlsx",
                        col_names = c("entry","name","icecream","pet", 
                                      "weekday","floors","yearborn"),
                        skip = 1)
```

## Writing spreadsheets
```{r Basic_R_IO-9}
writexl::write_xlsx(survey, "sample2.xlsx")
```

```{r Basic_R_IO-10, echo=FALSE, results='hide'}
file.remove("sample2.xlsx")
```

```{r Basic_R_IO-11}
openxlsx::write.xlsx(survey, "sample2.xlsx")
```

```{r Basic_R_IO-12, echo=FALSE, results='hide'}
file.remove("sample2.xlsx")
```

```{r Basic_R_IO-13}
openxlsx::write.xlsx(survey, "sample2.xlsx", sheetName="survey")
```

```{r Basic_R_IO-14, echo=FALSE, results='hide'}
file.remove("sample2.xlsx")
```

## Writing spreadsheets
```{r Basic_R_IO-15}
openxlsx::write.xlsx(survey, "sample2.xlsx", asTable=TRUE,
                     startCol=2, startRow=2)
```

```{r Basic_R_IO-16, echo=FALSE, results='hide'}
file.remove("sample2.xlsx")
```

## Exercises
1. Write a sample of 1000 rows from `nycflights13::flights` to excel, setting the column widths to automatically size (use the help text to find the argument name required)


# APIs

## APIs
APIs are parts of the web you can request (or provide) information from in a way that allows machines to talk to each other.

Each API request needs:

1. A URL
2. Any authentication info
3. Parameters / content

## Get data
```{r Basic_R_IO-17}
"https://haveibeenpwned.com/api/v2/breaches" %>% 
  httr::GET() %>% 
  httr::content(as = "text") %>% 
  jsonlite::fromJSON() ->
  breachedsites

head(breachedsites)
```

## Exercises
1. Go the HaveIBeenPwned API documentation and get the URL for retrieving a single breach
2. Use the URL to construct a request to find out if Snapchat have had a breach

# Databases
## Connections
```{r Basic_R_IO-18}
library(DBI)
library(odbc)
dbConn<-dbConnect(odbc(),
          driver="SQL Server",
          server="mhknbn2kdz.database.windows.net",
          database="AdventureWorks2012",
          uid="sqlfamily",
          pwd="sqlf@m1ly")

sample(dbListTables(dbConn),6)
```

## SQL
```{r Basic_R_IO-19}
betterTblsList<-dbGetQuery(dbConn, "SELECT * FROM information_schema.tables")
betterTblsList
```

```{r Basic_R_IO-20, eval=FALSE}
dbConn %>% 
  dbGetQuery("select * 
             from Production.TransactionHistory") ->
  transactions
```

```{r Basic_R_IO-21}
interestingProduct <- 797
dbConn %>% 
  sqlInterpolate("select * 
             from Production.TransactionHistory
             where productid = ?product",
             product = interestingProduct) %>% 
  dbGetQuery(dbConn, .) ->
  transactions797
```

## Table fetches
Currently limited to tables in default schema

```{r Basic_R_IO-22}
dbConn %>% 
  dbReadTable("SqlFamily") ->
  sqlfamily
```

## Exercises
1. Open the sampleSQL.r script and run the connection
2. Bring the four tables from the olympics schema into memory:
  + Medalist
  + Sport
  + Discipline
  + Event

## Limitations (currently)
- DBI has limited schema support
- DBI does not have an easy way to provide multiple values to `IN ()` clauses
- DBI doesn't support guids

## Writing SQL - JOINS
Common types of joins are:

- **inner join** for getting only rows matched on both sides
- **left join** for getting all the rows on the laft hand side with any matching data from the right
- **cross join** for getting every row on the left joined against every row on the right

##  Writing SQL - JOINS
```{r Basic_R_IO-23}
dbConn %>% 
  dbGetQuery(
    "select * 
    from northwind.orders o
    inner join northwind.[order details] d on o.orderid=d.orderid"
  ) ->
  orders
```

## Writing SQL - WHERE
Apply clauses that filter out records using the WHERE part of a SQL query. 

Use comparisons, build compound statements with AND and OR, and check for properties like whether a value is missing (IS NULL).


```{r Basic_R_IO-24}
dbConn %>%
  dbGetQuery(
    "select *
    from production.transactionhistory
    where productid > 800 
    and quantity > 1"
  ) ->
  limited_transactions
```

## Exercises
1. Write a query that joins all the olympics data together
2. Add a filter to the query that restricts the results to only the gold medal winners

## Writing queries - SELECT
- Refer to all columns with `*`
- Refer to all columns in a given table with `alias.*`
- Create new output columns or rename existing columns with `AS`
- Use `CASE WHEN` to create conditional columns

```{r Basic_R_IO-25}
dbConn %>% 
  dbGetQuery(
    "select o.*,
      d.unitprice * d.quantity AS TotalCost
    from northwind.orders o
    inner join northwind.[order details] d on o.orderid=d.orderid"
  ) ->
  orders_slim
```

## Exercises
1. Return a subset of columns from your joined up olympics data that you think are most relevant
