
```{r Basic_R_StringManipulation-1, include=FALSE, message=FALSE}
library(dplyr)
```

# General string tasks
## stringr
- All string manipulation functions prefixed with "str_".
- Takes the input string first (in contrast to base functions)

```{r Basic_R_StringManipulation-2, results='hide'}
library(stringr)
```

## Common tasks

```{r Basic_R_StringManipulation-3}
simple<-"This IS HOrrib1y typed! "
numbers<-c("02","11","10","1")
str_to_lower(simple)
str_trim(simple)
str_sort(numbers, numeric=TRUE)
str_length(numbers)
```

## Splitting
```{r Basic_R_StringManipulation-4}
str_split(simple,boundary("word"))
str_split(simple,boundary("sentence"))
str_split(simple,"i")
```

## Extraction
```{r Basic_R_StringManipulation-5}
str_detect(simple,"typed")
str_extract(simple,"typed.*")
```

## Counts
```{r Basic_R_StringManipulation-6}
str_count(simple, "i|I")
```

## Exercises
1. Store "the quick brown fox jumps over the lazy dog"
2. Upper case the string
3. Split the sentence into words and count the number of letters in each word

# Working with factors
## forcats
- All functions related to factors prefixed with "fct_"

```{r Basic_R_StringManipulation-7, results='hide'}
library(forcats)
```

## Common tasks
```{r Basic_R_StringManipulation-8}
myFactor<-as.factor(c("red","blue","yellow",NA,"red"))
fct_count(myFactor)
```

## Common tasks
```{r Basic_R_StringManipulation-9}
fct_explicit_na(myFactor) %>% 
  fct_count()
```

## Common tasks
```{r Basic_R_StringManipulation-10}
fct_infreq(myFactor) %>% 
  fct_count()
```

## Common tasks

```{r Basic_R_StringManipulation-11}
fct_lump(myFactor,n = 1) %>% 
  fct_count()
```

## Regulatory / security 
```{r Basic_R_StringManipulation-12}
fct_anon(myFactor) %>% 
  fct_count()
```

## Exercise
1. Count the records in the `gss_cat` dataset by survey year
1. Revise `gss_cat` such that there are only three levels of marital status
1. Race, partyid, & religion are sensitive values, anonymise them data

# String interpolation

## glue
- Clean dynamic string creation

```{r Basic_R_StringManipulation-13, results='hide'}
library(glue)
```

## Simple
```{r Basic_R_StringManipulation-14}
glue("string1", "string2", .sep=" ")

string2<-"was hard coded"
glue("string1 {string2}")
```

## More complicated
```{r Basic_R_StringManipulation-15}
glue("I went to the shop and I bought a {day1}",
     " and a {day2}",
     day1="ham", day2="cupcake")

glue("I purchased everything for £{round(cost, 2)}",
     cost=4/3)
```

## Complicated
```{r Basic_R_StringManipulation-16}
iris %>% 
  group_by(Species) %>% 
  glue_data("The average sepal width for {Species} is", 
            " {round(mean(Sepal.Width),1)}cm which is",
            " {round(mean(Sepal.Width),1)-Sepal.Width}cm bigger",
            " than this {Species}'s sepal width") ->
  results

results[sample(1:150, 5)]
```

## Exercise
1. Produce a statement that tells us what today's date is
2. Work out the average values for each species in iris and produce a sentence providing these values for each species
