---
title: "Data Manipulation"
author: "Steph Locke"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document: 
    code_folding: show
    number_sections: yes
    toc: yes
    toc_float: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Data Manipulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[Home](index.html) | [Prev](GettingStarted.html) | [Next](DataViz.html)

# Working with vectors {.tabset}
## Intro
Vectors are the building blocks of R. They are 1D arrays that contain data of all the same type - so coercion of data types happens.

- R is a 1-based array system so a vector starts at position 1
- Items in a vector can have names but can also be referred to by position

## Create
You can create vectors using the `c()` function or via a sequence (`:`). Values in a vector can be given names - this can be useful when subsetting values.
```{r}
handCrafted<-c(1,2,3,4)
seqCrafted<-1:4
named<-c(a=1,b=2,c=3,d=4)
named
```


## Filter
Alternatively called subsetting, we can filter a vector by using positive notations of position, negative notations of position, the name of a value, or providing a boolean value for each item in the vector.
```{r}
handCrafted[1]
seqCrafted[-1]
named["b"]

handCrafted[c(rep(TRUE,3),FALSE)]
```

## Update
You can update one or more values in a vector by assigning the new values into the desired subset.
```{r}
handCrafted
handCrafted[2]<-99
handCrafted

named
named["a"]<-99
named

# Delete by subsetting without value
seqCrafted
seqCrafted<-seqCrafted[-4]
seqCrafted

# Append by creating a vector combining the original and the additional values
ordered
ordered<-c(ordered,5)
ordered
```

## Manipulate
You can manipulate a vector using functions. You can overwrite any object created - this will change mode, class, etc as R is **loosely typed** and doesn't require such things to be specified and fixed up front.
```{r}
mode(seqCrafted)
seqCrafted<-as.character(seqCrafted)
mode(seqCrafted)
```

## Order
Ordering of records works by providing the position numbers of values in a vector and then using those to produce a vector with the original components in new locations
```{r}
preOrder<-sample(letters, 6)
preOrder

# Get order the values should appear in to be alphabetised
order(preOrder)

# Use it to sort a vector
ordered<-preOrder[order(preOrder)]
ordered

# Alternatively, use the sort() function for brevity
sorted<-sort(preOrder)
sorted
```

## Metadata
You can extract various pieces of information about a vector
```{r}
names(handCrafted)
names(named)

dim(named)
dimnames(named)

length(named)
class(named)
mode(named)

attributes(named)
str(named)
```

## Exercises
1. Create a vector containing upper case and lower case variants of the alphabet
1. Create a new vector with a random sample of your new letter vector
1. Filter out any lowercase letters

## Answers
```{r}
#1
lets<-c(LETTERS,letters)
#2
lets<-sample(lets,50)
#3
lets<-lets[tolower(lets)!=lets]
lets
```

# Working with lists {.tabset}
## Intro
Lists hold multiple objects together and form the basis of complex data objects like data.frames and model results.


## Create
You can create lists using the `list()` function or via a sequence (`:`). Objects in a list can be given names.
```{r}
basicList<-list(c(1,2,3,4),LETTERS[5:8], rnorm(5))
namedList<-list(p1=c(1,2,3,4),p2=LETTERS[5:8])
```


## Filter
Alternatively called subsetting, we can filter a list by using positive notations of position, negative notations of position, or the name of a element, or providing a boolean value for each item in the list. We can also use `list[[ "elementname" ]]` or `list$elementname for specifically detailing a *single element*. 
```{r error=TRUE}
basicList[1]
basicList[-1]
namedList["p2"]
basicList[c(TRUE,FALSE)]

basicList[[1]]
basicList[[-1]]
namedList[["p2"]]
namedList$p2
basicList[[c(TRUE,TRUE)]]
```

## Update
You can update one or more objects in a list by assigning the new values into the desired object using the same subsetting capabilities as noted in the Filter section.
```{r}
basicList[[1]]<-8:12
basicList[1]

# Elements in a list can be removed by making them NULL
basicList[2]
basicList[2]<-NULL
basicList[2]

# Append by creating a list combining the original and the additional values
basicList[[3]]<-LETTERS[5:8]
```

## Manipulate
You can manipulate a list using functions, and also manipulate the objects stored in the list.
```{r}
lapply(basicList,mode)
```

## Order
Ordering of objects in a list is rarely required but you can do it with the `order()` function.
```{r}
unorderedList<-list(p2=c(1,2,3,4),p1=LETTERS[5:8])
unorderedList[order(names(unorderedList))]
```

## Metadata
You can extract various pieces of information about a list
```{r}
names(basicList)
names(namedList)

dim(namedList)
dimnames(namedList)

length(namedList)
class(namedList)
mode(namedList)
attributes(namedList)
str(namedList)
```

## Exercise
Create a linear regression model (`lm()`) for the iris dataset and extract the `fitted.values` element

## Answers
```{r}
irisLM<-lm(Sepal.Width~Sepal.Length, iris)
head(irisLM$fitted.values)
```

# Working with tables {.tabset}
## Intro
Tables or data.frame's as the base structure is called in R can hold multiple columns of different data types. Normally, [data.table](https://cran.r-project.org/package=data.table) or [dplyr](https://cran.r-project.org/package=dplyr) would be taught to super-charge data.frames (see Steph's extended session "Cut the R learning curve" for more on these) but to reduce dependencies and make it easier to transfer code between different systems, we'll use just base R.

- data.frames have a coordinate system like excel so `df[ 1 , 2 ]` selects the intersection of row 1 and column 2
- Rows and columns can be referenced by name
- A data.frame is actuallly a list that is presented like a table

## Create
You can create data.frames using the `data.frame()` function. 

- `data.frame()` will error if you try storing something odd in it. Use `as.data.frame()` to coerce
```{r}
df<-data.frame(a=1:4, b=LETTERS[5:8], c=rnorm(4),row.names = letters[9:12])
df
```


## Filter
Alternatively called subsetting, we can filter a data.frame by using positive notations of position, negative notations of position, or the name of a element, or providing a boolean value for each item in the list. 

To filter using a condition requires the creation of a boolean vector based on a specific column. columns can be treated as vectors by using `df$colname`
```{r}
df[1, ]
df[ ,1]
df[1,1]
df[-(3:4),]
df[,"a"]
df[ , c(TRUE, TRUE, FALSE)]
df[df$a<4,]
```

## Update
You can update values in a data.frame by referencing it's position using the same subsetting capabilities as noted in the Filter section.
```{r}
df[1,1]<-2
df

# Columns in a data.frame can be removed by making them NULL
df[,2]
df[,2]<-NULL
df[,2]

# Rows in a data.frame can be removed by subsetting without them
df[2,]
df<-df[-2,]
df[2,]

# Appends can happen in a variety of ways
superDF<-data.frame(df,d=5:7)
superDF
df$newcol<-5:7
df
df[4,]<-c(1,1,1)
df
```
<!--

## Manipulate
You can manipulate a data.frames using functions, and also manipulate the columns in the data.frame
```{r}
lapply(basicList,mode)
```
-->

## Order
Ordering of data.frames works by providing the position numbers of row numbers in a vector and then using these to return data.frame rows in a specific order
```{r}
# Get order the values should appear in
order(df$c)

# Use it to sort a table
ordered<-df[order(df$c),]
ordered
```

## Metadata
You can extract various pieces of information about a list
```{r}
names(df)

dim(df)
dimnames(df)

length(df)
class(df)
mode(df)
attributes(df)
str(df)
```

## Exercises
1. Make a copy of the `iris` dataset to play with
1. Add a column with the estimated area of the sepals
1. Filter out any record with a petal width lower than average
1. Sort by species name in descending order

## Answers
```{r}
#1 
myIris<-iris
#2
myIris$Sepal.Area<-myIris$Sepal.Width * myIris$Sepal.Length
#3
avg<-mean(myIris$Petal.Width)
myIris<-myIris[myIris$Petal.Width>=avg,]
#4
myIris<-myIris[order(myIris$Species,decreasing = TRUE),]
head(myIris)
```
