---
title: "Calling R from SQL Server"
author: "Steph Locke"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document: 
    code_folding: show
    toc: yes
    toc_float: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Calling R from SQL Server}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[Home](index.html) | [Prev](SQLServerFromR.html) | [Next](AzureML.html)

## Executing R
Executing R from SQL Server requires the execution of a stored procedure called `sp_execute_external_script` that takes three main parameters:

- `@language` This gets set to R but isn't isn't it an interesting parameter with lots of possibilities for the future?
- `@script` This is the R code to be executed
- `@input_data_1` This is a SQL statement that will produce the starting dataset in R

Unfortunately, the return schema is not inherited by SQL Server so you (ideally) should provide an output specification  via `WITH RESULT SETS`
```{r}
library(RODBC)
#host<-"rmsftdp.northeurope.cloudapp.azure.com"
host<-"104.45.80.173"
db<-"AWDW16"
uname<-"Rdemo"
pwd<-"tugait"
connstring<-sprintf("DRIVER={ODBC Driver 11 for SQL Server};Server=%1$s;Database=%2$s;Uid=%3$s;Pwd=%4$s;",host,db,uname,pwd)

conn<-odbcDriverConnect(connstring)
# convenience function
sqlServer<-function(sql) sqlQuery(conn, sql)
```


```{r}
sqlServer(
"EXECUTE   sp_execute_external_script
                @language = N'R'
              , @script = N'OutputDataSet <- InputDataSet'
              , @input_data_1 = N'SELECT 1 as Col'
 WITH RESULT SETS ((col int not null))           "
)
```

## Overcoming SQL limitations
R can come in handy when you want to do something but it's really ugly in SQL to achieve. 

For instance, if I had to produce monthly sales figures per reseller with months as columns, I'd have to use a dynamic PIVOT effect in SQL. In R the complexity is much reduced because pivoting is really simple in R
```{r}
res<-sqlServer(
"
EXECUTE   sp_execute_external_script
                @language = N'R'
              , @script = N'
			  # My preferred tbl manip pkg
			  library(data.table)
			  setDT(InputDataSet)
			  # Pivot!		  
			  OutputDataSet <- dcast(
			  InputDataSet[
			  order(OrderDateKey)
			  ,.(MonthTotal=sum(SalesAmount) )
			  ,.(ResellerName,Month= paste0(CalendarYear,\"-\", MonthNumberOfYear)
			  ) ]
			  , ResellerName ~ Month)
			  # Need to add our headers as an extra row to save worrying about specifying them
			  cols<-as.list(colnames(OutputDataSet))
			  OutputDataSet<-rbindlist(list(cols,OutputDataSet), use.names=FALSE, fill=FALSE)
			  '
			  
              , @input_data_1 = N'select 
              r.ResellerName,
              f.OrderDateKey,
              d.CalendarYear,
              d.MonthNumberOfYear,
              f.SalesAmount
              from FactResellerSales f
              inner join DimReseller r on f.ResellerKey=r.ResellerKey
              inner join DimDate d on d.DateKey=f.OrderDateKey'
          
"
)

knitr::kable(head(res[,1:6]))
```

## Limitations
- Output schema not inherited
- Single table passthrough (both ways)
- Output must be a data.frame

## Useful links
[BOL](https://msdn.microsoft.com/en-us/library/mt604845.aspx)

## Exercise
1. Calculate daily order volumes and totals of SalesAmount from FactResellerSales in SQL
2. Pass these values into an R query and build a (`lm`) model that predicts daily SalesAmount from daily sales volumes. Return the coefficients.

## Answer
```{r}
res<-sqlServer(
"
EXECUTE   sp_execute_external_script
                @language = N'R'
              , @script = N'
			  mod<-lm(SalesAmount ~ Volume, InputDataSet)
			  OutputDataSet<-data.frame(coefficients(mod), names(coefficients(mod)))
			  ', @input_data_1 = N'select 
              sum(f.SalesAmount)  as SalesAmount,
              count(*) as Volume
              from FactResellerSales f
              inner join DimReseller r on f.ResellerKey=r.ResellerKey
              group by OrderDateKey'
          
"
)
res
```
