---
title: "Calling SQL Server from R"
author: "Steph Locke"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document: 
    code_folding: show
    toc: yes
    toc_float: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Calling SQL Server from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[Home](index.html) | [Prev](DataViz.html) | [Next](RFromSQLServer.html)


## Building a connection
- Need a driver name!
- Need credentials (you can do trusted AD connections too)
- Need to store the connection for use
```{r}
library(RODBC)
#host<-"rmsftdp.northeurope.cloudapp.azure.com"
host<-"104.45.80.173"
db<-"AWDW16"
uname<-"Rdemo"
pwd<-"tugait"
connstring<-sprintf("DRIVER={ODBC Driver 11 for SQL Server};Server=%1$s;Database=%2$s;Uid=%3$s;Pwd=%4$s;",host,db,uname,pwd)

conn<-odbcDriverConnect(connstring)
# Test!
if(conn!=-1L) knitr::kable(sqlTables(conn,schema = "dbo"))
```

## RODBC functions
RODBC comes with some built-in functions for working with your database
```{r eval=FALSE}
dt<- sqlFetch(conn, "DatabaseLog")
sqlCopy( conn, "select * from DatabaseLog", "DatabaseLog2")
sqlSave( conn, dt[1:100,], "DatabaseLog2")
sqlClear(conn, "DatabaseLog2")
sqlDrop( conn, "DatabaseLog2")
```

## SQL
Use normal SQL statements using `sqlQuery()` - must send one statement at a time to get the results back from each.

```{r}
dt<- sqlQuery(conn,"
         SELECT 
          DatabaseLogID
          ,DatabaseUser
          ,Event
          ,[Schema]
          ,Object
          FROM DatabaseLog
         ")
knitr::kable(head(dt))
```


```{r}
str( sqlQuery(conn,"
        SELECT COUNT(*) as Vol FROM DatabaseLog;
         SELECT DatabaseLogID FROM DatabaseLog;
         ")
)
```

## Work "at scale"
Microsoft R Server comes with an additional package `RevoScaleR` that allows you to connect to SQL Server and process very large amounts of data without loading it all into memory. This package is not available in R or Microsoft R Server Open.

Read more about it [Getting Started with RevoScaleR](https://msdn.microsoft.com/en-us/microsoft-r/rserver/rserver-scaler-getting-started)

## Exercise
1. Return all FactResellerSales data along with the associated Reseller details

## Answer
```{r}
sales<- sqlQuery(conn,"
        SELECT TOP 10 * FROM FactResellerSales f
        INNER JOIN DimReseller r
          ON f.resellerkey = r.resellerkey
         ")
knitr::kable(head(sales))
```
