# Shiny Modules

## What's a Shiny Module?
A reusable block of Shiny code that can be used in multiple places.

It must contain a server function and can include corresponding UI and Input functions.

Each module contains a name space so each instance generates it own unique input and output IDs.

## Example
```{r eval=FALSE}
library(shiny)
mod_sample_ui <- function(id){
  ns <- NS(id)
  tagList(
  p(id)
  )
}
mod_sample_server <- function(input, output, session){
  ns <- session$ns
  print("yo")
}
shinyApp(
  ui=fluidPage(
    mod_sample_ui("1st"),
    mod_sample_ui("2nd")),
  server=function(input, output, session){
    callModule(mod_sample_server,"1st")
    callModule(mod_sample_server,"2nd")
  }
)
```

## Shiny Module structure
### Contents
Building a module requires a few functions that will typicaly reside in the R directory of a package, in a file sourced by a shiny app, or included in a shiny app file.

- A "server" function accepting `input`,`output`, and `session` at a minimum
- A "UI" function accepting an `id`
- An "input" function accepting an `id`

As these are functions, they can be contained within a package structure and documented like any other function to provide additional help to Shiny app developers.

We recommend the [{golem}](https://github.com/ThinkR-open/golem) package for quickly creating a module starter file.

## Inputs
Shiny modules can consume  module specific inputs or be passed extra values via additional arguments to the module function. 

Inputs can be generated and used within a module. 

- These won't be accessible outside the module, by default.
- To correctly identify such inputs, wrap the labels in inputs and outputs in UI functions with `ns()`.

```{r eval=FALSE}
localinput_server = function(input, output, session){
  output$text = renderText(input$name)
}

localinput_ui = function(id){
  ns <- NS(id)
  tagList(textInput(ns("name"),"Insert name here"), p(id),
          textOutput(ns("text")))
}

shinyApp(
  ui=fluidPage(
    localinput_ui("local"),
    localinput_ui("local2")),
  server=function(input, output, session){
    callModule(localinput_server,"local")
    callModule(localinput_server,"local2")
  }
)
```

Here, we can see how a global variable needs to be passed to a module.

- Module server code by default changes `input$i` to `input$ns-i` so global inputs are not easily referenced.
- A common way to get around this is to create an extra argument to the server function and pass the input as a reactive variable into it.

```{r eval=FALSE}
globalinput_server = function(input, output, session, name){
  output$text = renderText(name())
}

globalinput_ui = function(id){
  ns <- NS(id)
  tagList(textOutput(ns("text")))
}

shinyApp(
  ui=fluidPage(
    textInput("name","Insert name here"),
    globalinput_ui("global")),
  server=function(input, output, session){
    callModule(globalinput_server,"global",
               name = reactive(input$name))
  }
)
```


## Outputs
The typical outputs in Shiny are supported in modules. They need to be referenced as being within the namespace inside the UI function i.e. `renderText(ns("textthingy"))`.

```{r eval=FALSE}
plotoutput_server = function(input, output, session){
  output$plot = plotOutput(
    ggplot2::qplot(x = Sepal.Width, y=Sepal.Length, data=iris)
  )
}

plotoutput_ui = function(id){
  ns <- NS(id)
  tagList(
    renderPlot(ns("plot"))
  )
}

shinyApp(
  ui=fluidPage(
    plotoutput_ui("plot")
    ),
  server=function(input, output, session){
    callModule(plotoutput_server,"plot")
  }
)
```



## Reactivity
Additionally, modules as functions can have outputs consumable inside the application. 

- These can be things like data.frames so that multiple dependant modules can rely on the same result instead of each doing the processing seperately.
- The outputs are typically reactive

```{r eval=FALSE}
sharedoutput_server = function(input, output, session){
  reactive(dplyr::sample_n(iris, input$n))
}

sharedoutput_ui = function(id){
  ns <- NS(id)
  tagList(numericInput(ns("n"), "n rows",value = 50))
}

output_server = function(input, output, session, shared){
  output$text = renderText(mean(shared()$Sepal.Width))
}

output_ui = function(id){
  ns <- NS(id)
  tagList(p(textOutput(ns("text"))))
}

shinyApp(
  ui=fluidPage(
    sharedoutput_ui("shared"),
    output_ui("one"),
    output_ui("two")),
  server=function(input, output, session){
    shared = callModule(sharedoutput_server,"shared")
    callModule(output_server,"one",shared=shared)
    callModule(output_server,"two",shared=shared)
  }
)
```

## Recommended reading
- [Original Modlarizing Shiny App Code post](https://shiny.rstudio.com/articles/modules.html)
- [{golem}](https://github.com/ThinkR-open/golem)
- [Shiny Module Design Patterns](https://github.com/stephlocke/shinymodulesdesignpatterns)