
```{r Basic_R_pivottabler-1, include=FALSE, message=FALSE}
library(pivottabler)
```




# pivottabler basics
## About
pivottabler was written by Chris Bailliss to make the production of pivot tables easier at Cancer Research UK.

There are substantial [vignettes](https://github.com/cbailiss/pivottabler/tree/master/vignettes) that extend the content beyond this basic training.

## Starting
To start producing a pivot table you need to create a pivot table object.

```{r Basic_R_pivottabler-2}
library(pivottabler)
pt <- PivotTable$new()
```

## Key methods
You can then add to your object via a number of commands:

- `addData` to attach a dataset
- `addColumnDataGroups` to put produce summaries by this field across the columns
- `addRowDataGroups` to put produce summaries by this field down the rows
- `defineCalculation` to determine the aggregation

## Using methods
These are not functions, but things an object can do so they get called via a `$` after referencing the object.

```{r Basic_R_pivottabler-3}
pt$addData(iris)
pt$addRowDataGroups("Species")
pt$defineCalculation(calculationName = "Mean",
                     summariseExpression = "mean(Sepal.Width)" )
pt$renderPivot()
```

## Easy pivots
With simple pivots an alternative can be used. There is a basic `qpvt`, a HTML rendering `qhpvt`, and a \LaTeX rendering `qlpvt`.

```{r Basic_R_pivottabler-4}
qhpvt(iris,"Species", calculations = "mean(Sepal.Width)")
```

## Groups
- We need to provide string representations of fields that should be put in either the rows or columns of a pivot table.
- Each additional command adds another level of depth to the rows or columns.
- On the fly calculations of categories are not supported

```{r Basic_R_pivottabler-5}
iris2<-iris; iris2$Small<-iris2$Sepal.Width<mean(iris2$Sepal.Width)
pt<-PivotTable$new()
pt$addData(iris2)
pt$addRowDataGroups("Species")
pt$addRowDataGroups("Small")
pt$renderPivot()
```

## Calculations
- We need to provide string representations of calculations that work in either dplyr e.g. `n()` or data.table e.g. `.N`
- Each additional command adds another calculation

```{r Basic_R_pivottabler-6}
pt$defineCalculation(calculationName = "Count", summariseExpression = "n()")
pt$defineCalculation(calculationName = "Max_Sepal.Width", summariseExpression = "max(Sepal.Width)")
pt$renderPivot()
```

## Exercises
1. Construct a quick pivot that gives the count of cars in `mtcars` by the number of cylinders
2. Construct a pivot that gives the average of the various sepal and petal attributes by species for `iris`

# Render formats
## Rendering
pivottabler is designed to make good looking output tables in a variety of output formats.

- `print()` to pretty print to console
- `saveHtml()` to produce a standalone HTML file
- `renderPivot()` to produce an inline HTML version
- `getLatex()` to get the \LaTeX code
- `asJSON()` to get a JSON representation (use `viewJSON()` to get an interactive HTML view of it)



## Exercises
1. Using `writeLines()` save a copy of the \LaTeX representation of one of your pivot tables to a ".tex" file
1. Using `writeLines()` save a copy of the JSON representation of one of your pivot tables to a ".json" file
1. Using pivottabler only, save a copy of the HTML representation of one of your pivot tables to a ".html" file

# Style

## Built-in themes
```{r Basic_R_pivottabler-7}
pt$theme <- "default"
pt$renderPivot()
```

## Built-in themes
```{r Basic_R_pivottabler-8}
pt$theme <- "largeplain"
pt$renderPivot()
```

## Built-in themes
```{r Basic_R_pivottabler-9}
pt$theme <- "compact"
pt$renderPivot()
```

## Making a theme
```{r Basic_R_pivottabler-10}
locke <- list(
  headerBackgroundColor = "rgb(33, 101, 182)",
  headerColor = "rgb(255, 255, 255)",
  cellBackgroundColor = "rgb(255, 255, 255)",
  cellColor = "rgb(77, 77, 77)")
theme <- getSimpleColoredTheme(pt, "Locke",colors=locke, fontName="Roboto, Arial")
pt$theme<-theme
pt$renderPivot()
```

## Adding specific styles
- To start styling a pivot table you need to create a pivot style object based off an existing pivot table.
- Use CSS attributes and values in a list
- Determine what cells to apply the style to
- Apply the style to the cells

## Adding specific styles
```{r Basic_R_pivottabler-11}
s <- PivotStyle$new(pt, styleName="cellHighlight", declarations=list("color"="red"))
g <- pt$getCells(specifyCellsAsList=TRUE, columnNumbers=2)
gc <- lapply(g, function(cell) {cell$style <- s})
pt$renderPivot()
```
## Adding specific styles
Identify cells to apply styles to with the following methods:

- `getCells()` for specifying by range / location
- `findCells()` for specifying by criteria
- `findColumnDataGroups()` for specifying column labels 
- `findRowDataGroups()` for specifying row labels 

## Conditional formatting
Can use `findCells()` with some conditions to apply a style

```{r Basic_R_pivottabler-12}
s <- PivotStyle$new(pt, styleName="cellHighlight", declarations=list("color"="blue"))
g <- pt$findCells(minValue=2, maxValue = 3.5)
gc <- lapply(g, function(cell) {cell$style <- s})
pt$renderPivot()
```

## Exercise
1. Add a conditional format to your average values by species where values are less than 3
