# rcpp

## Purpose
Run C++ code in R for faster calculations.

## Requirements
- `Rcpp` required
- `RcppArmadillo` recommended

```{r Basic_R_rcpp-1}
library(Rcpp)
```

## Basics

Create functions in-line with `cppFunction()`
```{r Basic_R_rcpp-2}
cppFunction(
"double cumulative_interest(double pv, float r, int p){
  double CI=pv*pow((1+r),p) - pv;
  return(CI);
            }")

cumulative_interest(100,.05,5)
```

A C++ function using `cppFunction()` requires:

- an output datatype
- a function name
- inputs with data types
- a body of code inside `{}`
    + code lines ended by `;`
    + a return statement

Functions with responses like `int` and `double` are not vectorised.

```{r Basic_R_rcpp-3, error=TRUE}
cumulative_interest(c(10,100),c(.1,.01),2)
```

To perform calculations that appear vectorised you need to provide vector inputs and outputs. Many operations have a vectorised capability so we don't have to write for loops.

```{r Basic_R_rcpp-4, error=TRUE}
cppFunction("DoubleVector  cumulative_interest2(DoubleVector pv, DoubleVector r, int p){
  return(pv*pow(1+r, p) - pv);
  }")

cumulative_interest2(c(10,100),c(.1,.01),2)
```

You can create files of C++ code and have anifty development experience in RStudio. You can then `sourceCpp()` or keep the fiels in a package.

## Exercise
1. Write a function that calculates the future value of an investment given an interest rate and a number of periods.

## Next steps
1. Read [Hadley's Rcpp tutorial](http://adv-r.had.co.nz/Rcpp.html)
2. Read [RStudio's page on Rcpp](https://support.rstudio.com/hc/en-us/articles/200486088-Using-Rcpp-with-RStudio)
3. Check out a recent [Rcpp workshop from Dirk](http://past.rinfinance.com/agenda/2017/workshop/DirkEddelbuettel.pdf)
