# Tidyeval

## Purpose
Tidyeval is the power behind the tidyverse. The tidyeval framework is primarily surfaced through the package `rlang`. The tidyeval package facilitates the making of functions, especially when working with the tidyverse.


## Requirements
- `rlang` required
- `tidyverse` recommended

```{r Basic_R_tidyeval-1}
library(rlang)
```

## Basics
`rlang` can handle the following:

- Stronger typed objects
- Extra object metadata
- Extra object assertions
- New function creation
- Non-standard evaluation
- Non-standed execution
- Managing environments
- Improved call stack and messaging

We will cover the fundamentals of everything bar environments and call stacks.

### Objects

We have new functions for making vectors of a specified data-type.

```{r Basic_R_tidyeval-2}
lgl(TRUE, 1)
int(TRUE, 1)
dbl(TRUE, 1, 1.1)
chr("a", "b")
```

Along with new functions for testing datatypes.
```{r Basic_R_tidyeval-3}
is_integer(lgl(TRUE, 1))
is_logical(int(TRUE, 1))
is_character(dbl(TRUE, 1, 1.1))
is_double(chr("a", "b"))
```

We can also perform easier metadata changes.
```{r Basic_R_tidyeval-4}
pi
pi %>% set_names("pi")
```

```{r Basic_R_tidyeval-5}
pi %>% set_attrs(cake=FALSE, names="pi")
```

There are also functions for simplifying list objects in various ways.
```{r Basic_R_tidyeval-6}
my_list<-list(letters[1:5],LETTERS[1:3],
              list(letters[10:15],
                   LETTERS[8:10]))
my_list
```

Remove a level
```{r Basic_R_tidyeval-7}
flatten(my_list)
```

Remove all levels
```{r Basic_R_tidyeval-8}
squash(my_list)
```

Specify type to get consolidation of vectors

```{r Basic_R_tidyeval-9}
rlang::flatten_chr(my_list)
```


#### Exercise
1. Create a new `letters` vector that is a character vector.
2. Add attributes saying which alphabet it is and how many letters are in it

### Functions

Build and assign a function to an environment with `new_function`.
```{r Basic_R_tidyeval-10}
calc<-new_function(list(a=1,b=2), quote(a*b))
calc()
calc(2,3)
```

Alternatively, convert *lambda functions* to full functions using `as_function`.

These lambda functions start with a tilde (`~`) and contain a calculation with up to 2 input arguments. Use `.` for just one input argument and `.x` and `.y` for two input arguments.

```{r Basic_R_tidyeval-11}
calc2<-as_function(~.x*.y)
calc2(2,3)
```

#### Exercise
1. Create a function that calculates the area of a circle givene a radius. Use the default value for pi, unless someone wants to provide a more precise version.
2. Create a function that converts from fahrenheit to celcius.

### Quasiquotation
- `quo` quote something
- `enquo` quote something or an expression
- `expr` quote something provided to a function argument
- `quo_name` get the name of something
- `!!` unquoute and evaluate  operator
- `!!!` unquote, combine, and evaluate operator

Use `quo` to make something context aware and `!!` to evaluate a context aware value.

```{r Basic_R_tidyeval-12}
library(dplyr)
my_var <- quo(homeworld)

starwars %>%
  group_by(!!my_var) %>%
  summarise_at(vars(height:mass), mean, na.rm = TRUE)
```

Use `enquo` to allow expressions.

```{r Basic_R_tidyeval-13}
summarise_starwars <-function(col){
  my_var <- enquo(col)

starwars %>%
  group_by(!!!my_var) %>%
  summarise_at(vars(height:mass), mean, na.rm = TRUE)
}

summarise_starwars(species)
```

```{r Basic_R_tidyeval-14}
summarise_starwars(substr(species,1,1))
```

```{r Basic_R_tidyeval-15, error=TRUE}
totals_by_col <-function(tbl, grp_col, dbl_col, na.rm=TRUE){
  grp_col <- enquo(grp_col)
  dbl_col <- enquo(dbl_col)
  dbl_nm <- quo_name(dbl_col)

  tbl %>%
    group_by(!! grp_col) %>% 
    summarise(!!dbl_nm := sum(!!dbl_col, na.rm = na.rm))
}

totals_by_col(starwars, gender, height)
```

### Exercise
1. Write a function that returns the count of unique values for a column when provided with a table name and a column name.

## Next steps
1. Continue learning to program with tidyeval in the [dplyr programming vignette](http://dplyr.tidyverse.org/articles/programming.html)
