# reticulate

## Purpose
Use `reticulate` to integrate python into your R workflows.

## Requirements
- `reticulate` package required
- A python installation required
- Python install in $PATH ideally

```{r Basic_R_reticulate-1}
library(reticulate)
py_discover_config()
```

## Basics
The `reticulate` package opens contexts to packages via `import()`.

```{r Basic_R_reticulate-2}
scipy <- import("scipy")
```

These contexts then make package functionality available as methods.

```{r Basic_R_reticulate-3}
scipy$amin(c(1,3,5,7))
```

Get access to the default python functionality with some extra functions.

```{r Basic_R_reticulate-4}
main <- import_main()
py <- import_builtins()
```

Create python objects and assign them to memory.
```{r Basic_R_reticulate-5}
results <- scipy$amin(c(1,3,5,7))
``` 

Note that you currently cannot translate between pandas data.frames and R data.frames without utilising an intermediate write to disk step.

## Tensorflow
Tensorflow, a deep learning package in python, has been wrapped in R.

```{r Basic_R_reticulate-6, eval=FALSE}
library(tensorflow)
#install_tensorflow()
sess = tf$Session()
hello <- tf$constant('Hello, TensorFlow!')
sess$run(hello)
```

Check out the [mnist tutorial](https://tensorflow.rstudio.com/tensorflow/articles/tutorial_mnist_beginners.html) to see a more indepth example of running tensorflow via R.

## Exercise
1. Import the python package requests
2. Make a request using the get function and extract the content attribute

```{r Basic_R_reticulate-7, echo=FALSE, results="hide"}
c <- import("requests")
c$get("https://google.co.uk")$content
```
## Next steps
1. Read the [docs](https://rstudio.github.io/reticulate/index.html)
