# golem

`golem` is a package designed to help people develop bigger shiny applications. If provides a template shiny app and uses a package structure to maintain dependencies and shiny code in **modules**.

```{r eval=FALSE}
remotes::install_github("ThinkR-open/golem")
```


```{r eval=FALSE, tidy=FALSE}
golem::fill_desc(
  pkg_name = , # The Name of the package containing the App 
  pkg_title = , # The Title of the package containing the App 
  pkg_description = , # The Description of the package containing the App 
  author_first_name = , # Your First Name
  author_last_name = , # Your Last Name
  author_email = , # Your Email
  repo_url = NULL) # The (optional) URL of the GitHub Repo

golem::use_recommended_dep()
```

Structuring the application can be done with `golem` functions:

```{r eval=FALSE}
golem::use_utils_ui()
golem::use_utils_server()
golem::add_module(name = "this")
golem::use_recommended_tests()
```

## Exercise
Create a shiny project using golem. Structure it with a server and a ui component and present the iris data with a species filter in the application.