
# dbplyr

dbplyr allows us to open connections to database objects, write dplyr and have the database execute the SQL equivalent.

- `dbplyr` required
- `dplyr` required
- `DBI` required
- `odbc` recommended
- Full `tidyverse` recommended
- A database required

## Basics

```{r Basic_R_dbplyr-1}
library(DBI)
library(odbc)

driver   = "ODBC Driver 17 for SQL Server"
server = "lockedata.database.windows.net"
database = "datasci"
uid = "lockedata"
pwd = "zll+.?=g8JA11111"

dbConn<-dbConnect(odbc(),
          driver=driver, server=server,
          database=database, uid=uid,
          pwd=pwd)
```


With a database connection we can build connections to individual tables using `tbl()`.

```{r Basic_R_dbplyr-2}
library(tidyverse)
library(dbplyr)

if(!("flights" %in% dbListTables(dbConn))){
copy_to(dbConn, nycflights13::flights, "flights",
  temporary = FALSE, 
  indexes = list(
    c("year", "month", "day"), 
    "carrier", 
    "tailnum",
    "dest"
  )
)
}

flights<-tbl(dbConn,"flights")
```

With these table references as our starting points, we can then write queries against them like they were in-memory R data.frames. These get translated in to SQL, if possible, and executed against the database, saving computation and memory headaches on the R side.

```{r Basic_R_dbplyr-3}
flights %>% 
  filter(Month==1) %>% 
  summarise(n())
```

By default, R preps things until you force it to execute the code against the DB by displaying the results. Sometimes you need to force a collection of the data by using `collect()` so that you can use functionality that only works on data.frames, like `ggplot()`.


```{r Basic_R_dbplyr-4}
flights %>% 
  filter(Month==2) %>% 
  summarise(n()) %>% 
  collect()->
  feb_count

feb_count
```

## Debugging
Not every R function or transformation is compatible with your database so you will sometimes get errors that mean you will need to change your code -- either by changing the R functions used, or by collecting data first. 

I would also not recommend you run dbplyr code as part of a production data pipeline. It will not generate code as optimised as a DBA who knows the DB could write it so it should be used for formalising requirements or ad-hoc work.

To get the SQL for debugging or for optimisation, you can use `show_query()`.

```{r Basic_R_dbplyr-5}
flights %>% 
  filter(Month==1) %>% 
  summarise(n()) %>% 
  show_query()
```

## Joins
If you have multiple table connections within a database, you can write joins between the tables and have this translate to SQL that will execute in the database.

```{r Basic_R_dbplyr-6}
if(!("flights_carriers" %in% dbListTables(dbConn))){
  nycflights13::flights %>% 
    count(carrier) %>% 
    slice(1:10) %>% 
    copy_to(dbConn, ., "flights_carriers", 
  temporary = FALSE)
}

carriers<-tbl(dbConn, "flights_carriers")

flights %>% 
  inner_join(carriers, by=c("carrier"="carrier")) %>% 
  show_query()
```

## Exercise
1. Find the average departure delay (assume NAs are 0 delay) by destination from the flights table
2. Find the count of flights with carriers not in the flights_carriers table

## Next steps
1. Read about how dplyr gets mapped to SQL at [dbplyr.tidyverse.org](http://dbplyr.tidyverse.org)
2. Extend the [MSSQL conversions](https://github.com/tidyverse/dbplyr/blob/master/R/db-odbc-mssql.R)
