# V8

## Purpose
Use V8 to execute ad-hoc javascript or leverage pre-existing code in javascript.

## Requirements
- `V8` package required
- [V8 system dependencies](https://github.com/jeroen/V8#installation) on Linux

```{r Basic_R_V8-1}
library(V8)
```

## Basics
To run js code, you need a context.

```{r Basic_R_V8-2}
js<-v8()
```

We can send javascript code to be executed using the `eval` method.

```{r Basic_R_V8-3}
js$eval("Date()")
```

We can create variables inside our js context for use later.
```{r Basic_R_V8-4}
js$eval("var today= new Date()")
js$eval("var val=3+5")
js$eval("today")
js$eval("Math.pow(val,2)")
```

We can send data to our JS context using the `assign` method.

```{r Basic_R_V8-5}
js$assign("iris", iris)
js$eval("iris.length")
```

We can retrieve data using the `get` method.
```{r Basic_R_V8-6}
head(js$get("iris"))
```

We're not limited to native javascript functionality. We can add code from other sources like js libraries using the `source` method.

```{r Basic_R_V8-7}
js$source("https://raw.githubusercontent.com/MangoTheCat/swli/master/inst/js/forcem.js")
```

We can use functions using the `call` method
```{r Basic_R_V8-8}
# fn, episode, paragraphs
js$call("forcem", "e4", 1)
```

## Exercise
1. Play with the Lorem Ipsum generator.

## Next steps
1. Read the [docs](https://cran.r-project.org/package=V8)
2. Check out [swli](https://github.com/MangoTheCat/swli)
