# Decision trees

## Decision trees
Purpose: predict a discrete outcome


```{r Basic_Models_DecisionTrees-1, echo=FALSE}
titanic %>% 
  FFTrees(survived ~., data=.) %>% 
  plot(main = "Titanic", 
     decision.names = c("Died", "Survived"))
```

## Function 
Therer are number of decision tree R packages, here I use FFTrees for the superior visualisations.

```{r Basic_Models_DecisionTrees-2}
titanic %>% 
  FFTrees(survived ~., .)
```

## Getting an overview

```{r Basic_Models_DecisionTrees-3}
titanic %>% 
  FFTrees(survived ~., .) %>% 
  summary()
```

## Getting a visual overview

```{r Basic_Models_DecisionTrees-4}
titanic %>% 
  FFTrees(survived ~., .) %>% 
  plot()
```


## Getting variable importance

```{r Basic_Models_DecisionTrees-5}
titanic %>% 
  FFTrees(survived ~., .) %>% 
  plot(what="cues")
```

## Making predictions

```{r Basic_Models_DecisionTrees-6}
titanic %>% 
  sample_n(10) ->
  testdf

titanic %>% 
  FFTrees(survived ~., .) %>% 
  predict(.,data=testdf) %>% 
  plot()
```

## Exercises
1. Build a decision tree for predicting the `y` column in the `bank` dataset
2. What are the most important factors in this tree?
