# (PART) Data Science Fundamentals {-} 
# Initial exploration
You should explore your data to look for issues and get a general understanding of the data.

```{r Basic_ML_DataPrep-1}
library(tidyverse)
library(AppliedPredictiveModeling)
data(AlzheimerDisease)
predictors %>% 
  cbind(diagnosis) ->
  alzheimers
```

Common things to explore/check for:

- Unusual distributions
- Outliers
- Missings
- Correlations


We can get an effective view of our data using the `skimr` package.

```{r Basic_ML_DataPrep-2}
alzheimers %>% 
  skimr::skim_to_wide()%>% 
  head()
```


Visualise the relationships between numeric variables using the original `pairs()` function or `ggpairs()` from `GGally`. This overview visualisation method, is good for up to maybe twenty variables before it's really difficult to a) compute and b) visualise.

```{r Basic_ML_DataPrep-3, fig.height=10}
alzheimers %>% 
  select_if(is.numeric) %>% 
  select(1:10) %>% 
  GGally::ggpairs(aes(alpha=.1))
```

Alternatively, the `xray` package can be useful, especially with lots of variables for looking at distributions.

```{r Basic_ML_DataPrep-4, fig.height=15}
alzheimers%>% 
  select(1:10) %>% 
  xray::distributions()
```

It can also be used to get a quick view of potential anomalies.
```{r Basic_ML_DataPrep-5}
alzheimers %>% 
  xray::anomalies()%>% 
  head()
```

If you have data which includes a date, you can use the `timebased()` function to generate graphs for visually inspecting if distributions changed over time.

There are a lots of other useful packages out there for exploring your data. The`visdat` package gives us some options for exploring our dataset at a high level.
```{r Basic_ML_DataPrep-6}
alzheimers %>% 
  visdat::vis_dat()
```

```{r Basic_ML_DataPrep-7}
alzheimers %>% 
  visdat::vis_miss(cluster=TRUE)
```

## Automated and assisted EDA
Generating lots of charts etc can be a pain. There are somee tools that can help you. The `IEDA` package provides an interactive exploratory data analysis shiny application that you can use to explore your data. 

Alternatively, or as well as, you can use the DataExplorer package and it's `create_report()` function to generate a good quality report overviewing the data.

```{r Basic_ML_DataPrep-8, eval=FALSE}
DataExplorer::create_report(alzheimers,y = "diagnosis")
```

## Exercise
1. Load the `titanic3` dataset from the `PASWR` package
3. Summarise the data 
4. If we're trying to predict who is going to die on the titanic, which column(s) can we immediately remove from our dataset?




# Cleaning data
We can use `janitor` to do some preparation and basic cleanup.

```{r Basic_ML_DataPrep-9}
head(colnames(alzheimers),20)

alzheimers %>% 
  janitor::clean_names() ->
  alzheimers


head(colnames(alzheimers),20)
```

Other cleanup steps will often involve manipulation of values, based on the data itself. `lubridate`, `stringr`, and `forcats` will often come in handy. `recipes` can then be used for feature preparation and reduction.


