# (PART) Core models {-}
# Linear regression
```{r Basic_Models_LinearRegressions-1, include=FALSE}
library(tidyverse)
library(modelr)
library(ggplot2)
library(broom)
library(FFTrees)
```


## Linear regression
Purpose: perform a line of best fit analysis to predict a continuous variable

- Works best on normally distributed outcomes and input data
- Fast computation
- Readily interpretable
- Needs work to handle non-linear relationships

A *linear regression* predicts a continuous variable based on multiple independent variables. 


If we were trying to predict a continuous variable:

```{r rnorm2, fig.cap="Continuous outcome variable"}
library(ggplot2)
set.seed(777)
y_n<-rnorm(1000,100,25)
qplot(y_n, binwidth=5)+theme_minimal()
```

A basic regression using a single variable could look like:
```{r xyc1, fig.cap="Simple line of best fit"}
x_n<-y_n+rnorm(1000,30,20)
qplot(x_n,y_n) + geom_smooth(method = "lm", se = FALSE)+theme_minimal()
```

## Function
The `lm()` or linear model function is used to calculate a linear regression model. The default display shows the model coefficients.

```{r Basic_Models_LinearRegressions-2}
lm(Sepal.Length~Petal.Width, data = iris)
```

The data argument comes second so when we want to use the tidyverse, we need to use our dot notation.
```{r Basic_Models_LinearRegressions-3}
iris %>% 
  lm(Sepal.Length~Petal.Width, .)
```

## Summarising the model
To get a more in-depth view of our model we can use `summary()`.
```{r Basic_Models_LinearRegressions-4}
iris %>% 
  lm(Sepal.Length~Petal.Width, .) %>% 
  summary()
```

## Extracting coefficients
To get the coefficients for use in other systems, plus permformance measures use broom's `tidy()`.
```{r Basic_Models_LinearRegressions-5}
iris %>% 
  lm(Sepal.Length~Petal.Width, .) %>% 
  broom::tidy()
```

## Extracting fitted values
We can get the original data plus the fitted values and error measurements out using broom's `augment()`.

```{r Basic_Models_LinearRegressions-6}
iris %>% 
  lm(Sepal.Length~Petal.Width, .) %>% 
  broom::augment() %>% 
  ggplot(aes(x=Sepal.Length, y=.fitted), data=.) +
  geom_point() +
  geom_abline(slope=1, intercept=0)
```

## Predictions

Make predictions and add theese to our original data with modelr's `add_predictions()` function.

```{r Basic_Models_LinearRegressions-7}
iris %>% 
  lm(Sepal.Length~Petal.Width, .) ->
  irismodel

iris %>% 
  sample_frac(.3) %>% 
  modelr::add_predictions(irismodel)
```

## Model summary statistics
To get overall quality measures about a model use broom's `glance()`.
```{r Basic_Models_LinearRegressions-8}
iris %>% 
  lm(Sepal.Length~Petal.Width, .) %>% 
  broom::glance()
```

## Exercises
1. Using the dataset airquality, build some models predicting the month based on weather attributes
2. Using AIC as your selection criteria, which of your models performed best?
